---
title: "Stat 579 Project"
author: "Andrew Maloney"
date: "10/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
# Can use this function to quickly install and load all packages

#Load required Packages



#packages <- c("readxl", "tidyverse", "purrr", "shiny", "shinydashboard",
             # "shinydashboardPlus", "shinythemes", "stringr", "ggplot2", "plotly", "lubridate",
              #"viridis", "rayshader")

#for (pkgs in packages){
# if(!require(pkgs, character.only = TRUE)){ # Condition 
#  install.packages(pkgs) # Install if not in set 1
# library(pkgs) # Load if installed
#}
#}





```


```{r, warning=FALSE}


library(readxl)
library(tidyverse)
library(purrr) #optional
library(shiny) #optional
library(shinydashboard) #optional
library(shinydashboardPlus) #optional
library(shinythemes) #optional
library(stringr)
library(ggplot2)
library(plotly)
library(lubridate)
library(viridis) #optional
library(rayshader) #optional (Heike hates this package but I love it lol)

library(naniar)

library(tidyr)

```


```{r, warning=FALSE}

df <- read_excel("all_lawsuits_2008_to_2019.xlsx")

head(df)
```

#### Case Number: Identification of the case
#### Payee: Person or group of people who is/are owed money 
#### Payment Amount:  Amount paid to the Payee
#### Fees and Costs: Costs associated with the settlement
#### Primary Cause: Short insight behind the settlement
#### City Department Involved: Department of Chicago involved in the settlement
#### Disposition: Final type of Settlement
#### Tort: Number of wrongful acts or infringement of rights leading to civil legal liability(big no no)
#### Date to Comptroller: Date to comptroller ?
#### Year to Comptroller: Year to Comptroller ?
#### Monto to Comptroller: Month to Comptroller?
#### Payment Amount(Millions): Payment amount in terms of millions of dollars
#### Fees and Costs(Millions): Fees and Costs in terms of millions of dollars
#### Total Paid: Payment Amount + Fees and Cost
#### Total Paid(millions): Total Paid in terms of millions of dollars

```{r}
#Spaces in column names are a no no for me so I am going to replace them with underscores using gsub

names(df) <- gsub(" ", "_", names(df))

```


```{r}
# Let's first investigate the missing values throughout the data set


vis_miss(df)

df %>%
  gather(key = "columns", value = "column_values") %>%
    mutate(missing_values = is.na(column_values)) %>%
      group_by(columns) %>%
        mutate(total_values = n()) %>%
          group_by(columns, missing_values, total_values)%>%
       summarise(number_of_missing_values = n()) %>%
        #filter(missing_values == TRUE) %>% select(-missing_values) %>% can see total number of NA values for each column
          mutate(percent_missing = (number_of_missing_values/total_values) * 100) %>%
  ggplot(aes(x = reorder(columns, desc(percent_missing)), y = percent_missing, fill = missing_values))+
    geom_bar(stat = "identity")+
      coord_flip()+ 
        ggtitle("Percentage of Missing Values throughout the Dataset")+
          xlab("Column Names")+ylab("Percentage")+
            theme_bw() + scale_fill_manual(values = c("Chartreuse3", "firebrick2"))


# 0.48% missing values for Case_Number
# <0.1% missing values for Payment_Amount
# <0.1% missing values for Fees_and_Costs
# <0.1% missing values for Primary_Cause
# 50.36% missing values for Tort
# <0.1% missing values for Date_to_Comptroller
# <0.1% missing values for Year_to_Comptroller
# <0.1% missing values for Month_to_Comptroller
# <0.1% missing values for Payment_Amount(millions)
# <0.1% missing values for Fees_and_Costs(millions)
# <0.1% missing values for Total_Paid
# <0.1% missing values for Total_Paid(Millsions)
# 100 % missing values for Payment_Fund, Effective_Data, Due_Date, and Client_Department

# 3 Columns contain no missing values
# 16 columns contain missing values, 4 out of those 16 columns are completely missing

```



```{r}
# Columns that are 100% missing data are of no use so I am removing them

df <- df %>% select(-Payment_Fund, -Effective_Date, -Due_Date, -Client_Department)

```

## Analysis of remaing missing values

```{r}

#Tort column is easy fix, NA should be 0

df <- df %>%
  mutate(Tort = ifelse(is.na(Tort) == TRUE, 0, Tort))


```

```{r}

df %>%
  gather(key = "columns", value = "column_values") %>%
    mutate(missing_values = is.na(column_values)) %>%
          group_by(columns, missing_values)%>%
       summarise(number_of_missing_values = n()) %>%
        filter(missing_values == TRUE) %>% select(-missing_values) 


# There are only 79 missing values left.  Do these values have a pattern or are they just missing

df %>% filter(is.na(Case_Number) == TRUE) #Looks like case Numbers are missing on their own (1)

df %>% filter(is.na(Date_to_Comptroller) == TRUE) # Date_to_Comptroller is the joined Year, Month, including day, they each have 2 missing values making a total of 6.  This should be re-separated to include day names could be better as well. (3)

df %>% filter(is.na(Fees_and_Costs) == TRUE) # Fees_and_cost and Fees_and_costs(millions) are linked.  Can remove missing values or take average value based on Primary_Cause (2)

df %>% filter(is.na(Payment_Amount) == TRUE) # Payment_Amount and Payment_Amount(millions) are linked.  The Fees_and_Costs for this observation are really high, could be an entry error.  Need to look at average Payment_Amount and Fees_Costs (2) associated with "UNLAWFUL SEARCH/FALSE ARREST

df %>% filter(is.na(Total_Paid) == TRUE) # One of the observations for the row containing missing Total_Paid has a valid Payment amount. Possible Entry issue. The other one is linked with the really high Fees_and_Costs observation.  Therefore it could be possible that 3 observations are linked across Total_Paid, Payment_Amount, and Fees_and_Costs (2)

df %>% filter(is.na(Primary_Cause) == TRUE) # Missing primary cost deals with Transportation department and has a date.  Probably can look up or remove

```

# Marie's stuff begins here

## Are there specific times during the year where certain departments are involved in more lawsuits?

```{r, echo = T}
df2 <- df %>% group_by(City_Department_Involved) %>% summarize(dept_count = n()) %>% arrange(desc(dept_count))
```

There are 75 different city departments that had lawsuits filed against them. Many of the department names are very similar, e.g. water management/water and water management. These departments could be potentially combined for analysis purposes, but for this preliminary analysis, we will focus on departments that had more than 1,000 lawsuits filed against them. They are:

* Police (5,049 lawsuits)

* City wide (1428 lawsuits)

* Streets and sanitation (1419 lawsuits)

* Transportation (1297 lawsuits)

* Water management/water (1205 lawsuits)

We will use a visualization to see if these departments are more likely to be sued at certain times of the year. We will examine the data at the month level.

```{r, echo = T}
df_top5 <- df %>% filter(City_Department_Involved %in% c("POLICE", "CITY WIDE", "STREETS & SANITATION", "TRANSPORTATION", "WATER MGMT / WATER"))
df_top5 <- df_top5 %>% mutate(month = factor(Month_to_Comptroller))
df_top5 %>% ggplot(aes(x = Month_to_Comptroller, fill = City_Department_Involved))+
  geom_bar()+
  scale_x_continuous(breaks = seq(1, 12, 1))
```

We can see that the City Wide department was sued a lot during June. We can also see that the Police department tends to have the most lawsuits filed against it in every other month. Water Management/Water sees an uptick in lawsuits during April and May and again in September and October.